\url{https://erikbern.com/2014/11/29/deep-learning-for-chess.html}

In chess there are 64 squares and 12 piece types.

\begin{definition}
  A {\it position} comprises two bits of information:
  \begin{enumerate}
  \item An assignment to each of the 64 squares of one of the 12 piece types, or EMPTY.
  \item Whether white or black is to play next.
  \end{enumerate}
  A {\it valid position} is a position for which the counts of the piece types are less than or equal to their initial counts.

  A {\it terminal position} is a valid position which is a win for white, a win for black, or a draw.
\end{definition}
Define the set $F = \{-1, 0, 1\}$ of labels. We label a terminal position $p$ as follows:
\begin{align*}
  f(p) =
  \begin{cases}
  -1 & \text{if $p$ is a win for white} \\
  0  & \text{if $p$ is a draw} \\
  +1 & \text{if $p$ is a win for black}.
  \end{cases}
\end{align*}
We extend this definition recursively to non-terminal positions as follows. Define $M(p)$ to be the set of
valid positions reachable in one move from $p$. The label at a non-terminal position $p$ is
\begin{align*}
  f(p) = \min_{p' \in M(p)} f(p').
\end{align*}
Thus, for

We now want to extend this definition to assign labels $f(p) \in F$ to non-terminal positions.

We do so (conceptually) according to the following algorithm:

\begin{definition*}
  \begin{itemize}
  \item Define $M(p)$ to be the set of valid positions reachable in one move from $p$.
  \item Define $M^{-1}(p)$ to be the set of valid positions from which $p$ is reachable in one move.
  \end{itemize}
\end{definition*}

#+begin_src python
def visit(position):
    if position.is_terminal:
        return

    ancestors = position.get_ancestors()

    for


#+end_src

\begin{enumerate}
\item Let $T_w$ be the set of all terminal positions (white to play).
\item For all $$$q \in T_w$
\end{enumerate}

Let $p$ (white to play) be a valid non-terminal position, and let $M(p)$ be the set of valid positions reachable in one move from $p$.

Note that if some $q = -1$ for $q \in M(p)$ then white can win in one move. Accordingly, we define $f(p) = -1$ if $-1 \in M(p)$.

Alternatively, suppose that $-1 \not\in M(p)$.

Let $q$ be a terminal position (black to play) and consider the set $P$ of positions (white to play) from which $q$ is reachable in one move.

Note that if $q = -1$ then white can win in one move from all positions $p \in P$. Therefore we will define $p = -1$ for all .$p \in P$.






Define $f(p) \in \{-1, 0, 1\}$ as follows:

Assuming that
