\begin{question*}
  We have $n$ data points $(x_{1}, y_{1}), \ldots, (x_{2}, y_{2})$. Find $y(x)$ which minimizes the sum of squares.
\end{question*}

How do we represent this as an integral of a differentiable Lagrangian?

Under the conventional least-squares criterion, the ``action'' is
\begin{align*}
  A[y] = \sum_{i=1}^{n} \(y(x_{i}) - y_{i}\)^{2}.
\end{align*}
We need to represent this as
\begin{align*}
  A[y] = \int_{y} \Lag(y, y', x),
\end{align*}
or
\begin{align*}
  A[y] = \int_{x_1}^{x_n} \Lag\(y(x), y'(x), x\) \dx.
\end{align*}
One simple choice of lagrangian would be to compute, (at every point along the candidate curve) the sum of
squared distances to all data points:
\begin{align*}
  \Lag(x, y) = \sum_i (x - x_i)^2 + (y - y_i)^2.
\end{align*}
The derivatives are
\begin{align*}
  \pdLdy &= \sum_i 2(y - y_i) = 2ny - 2\sum_i y_i \\
  \pdLdyp &= 0.
\end{align*}
The Lagrange equations are then
\begin{align*}
  \ddx \pdLdyp = \pdLdy,
\end{align*}
so we have that $\pdLdy = 0$ for all $x$ and therefore
\begin{align*}
  y = \frac{1}{n}\sum_i y_i
\end{align*}
for all $x$.


So, that's not what we wanted. This fitted line ignores the x-coordinates. Why?
\begin{itemize}
\item Is the argument correct and that's just what this choice of Lagrangian results in?
\item Or is the argument incorrect?
\item For $x$ to be involved (to get a non-trivial answer) it seems that we need a Lagrangian that has a dependency on $y'$.
\end{itemize}

I think


  Let's try a different Lagrangian: consider a point $(x, y)$ on the candidate curve, and consider data
  point $(x_i, y_i)$. If $x_i$ is close to $x$ but $y_i$ is far from $y$, then the Lagrangian should penalize
  the candidate curve. So define
  \begin{align*}
    \Lag(x, y) = \sum_i \frac{(y - y_i)^2}{1 + (x - x_i)^2}.
  \end{align*}
The partial derivatives are
\begin{align*}
  \pdLdy &=
\end{align*}
