\documentclass[12pt]{article}
\usepackage{mathematics}


\begin{document}

\section*{The relational model}
Consider a database table with $d$ columns.

In relational algebra, a column corresponds to a \emph{type} (aka \emph{domain}). A type is a set
of possible values.

A \emph{relation} (aka \emph{table}) is a subset of the Cartesian product of the $d$ domains.

Each element of a relation is a \emph{tuple} (aka \emph{row}).

Which tuples are in the relation is determined by a \emph{predicate} on the
relation (aka a constraint).

A \emph{database} is a collection of relations.


\section*{Rules Engine}

Let $X$ be a table of interest.

\begin{definition*}[value field]~\\
  Let $A$ be a type. In practice, a \textbf{value field} is the name of a
  column of type $A$ in a {\tt Rule} table. Conceptually, a \textbf{value
    field} is a function $X \to A$.

  Note that the value field is not a column of $X$.
\end{definition*}


\begin{example*}
  If the objective is to compute price for each row of $X$ according to a
  collection of rules, then the {\tt value\_field} might be {\tt "price"} where
  {\tt price} is a column of type Decimal in a {\tt Rule} table.
\end{example*}


\begin{example*}
  If the objective is to determine the subset of rows of $X$ that are flagged
  by a rule, then the {\tt value\_field} might be {\tt "flag"}, where {\tt
    flag} is a column of type Boolean in a {\tt Rule} table.
\end{example*}


\begin{definition*}[{\tt get\_value\_for\_instance}]~\\
  {\tt get\_value\_for\_instance} is a function\\
  $(x, \tt{value\_field}) \mapsto$ (the computed value for $x$ according to the
  rules).
\end{definition*}


\begin{definition*}[{\tt get\_preimage\_queryset}]~\\
  Let $k$ be a value field of type $A$ and let $v \in A$.

  Define $V:X \to A$ by $V(x) := {\tt get\_value\_for\_instance}(x, k)$.

  Then ${\tt get\_preimage\_queryset}$ is a function $A \to {\tt powerset}(X)$ defined by\\
  \begin{align*}
  {\tt get\_preimage\_queryset}(k, v) := V^\1(v).
  \end{align*}


  In other words, ${\tt get\_preimage\_queryset}(k, v)$ is the set of rows of $X$ for which the
  rules compute the value $v$ for value field $k$.
\end{definition*}



Our initial objective is to describe a rule selecting a subset of the rows of
$X$.



\end{document}