\documentclass[12pt]{article}
\usepackage{mathematics}

\newcommand{\true}{\texttt{true}}
\newcommand{\false}{\texttt{false}}

\begin{document}

\section*{The problem}
There are $m$ members of a club. Member $i$ has $f_i$ friends who are not yet in the club. There
are two relevant actions:
\begin{enumerate}
\item A member may {\it message} a friend to suggest joining the club.
\item Once messaged, the friend may {\it accept} the suggestion (join the club).
\end{enumerate}
The problem is to maximize the number of club members by suggesting friends to
message. Specifically, for each member we suggest $r$ of their friends for them to message.

\section*{Training data}

The training data contains a total of $n$ friends of $m$ members, i.e. $n = \sum_{i=1}^m f_i$.

For each friend we have $d$ features, so the training data $X$ is an $n \times d$ table.

There are two $y$ columns:
\begin{enumerate}
\item $y_1$ is $\true$ if friend $i$ was {\it messaged} by their associated member, else $\false$.
\item $y_2$ is $\true$ if friend $i$ was {\it messaged} by their associated member {\it and
    accepted}, else $\false$.
\end{enumerate}


\newpage
\section*{Model comparison}

\begin{tabular}{c|c|c|c|}
  &             &Predict invited  &Predict not invited\\
  \hline
  &Invited      &TP               &FN\\
  &Not invited  &FP               &TN
\end{tabular}

\begin{itemize}
\item TN rate = 1 - FP rate
\item TP rate = 1 - FN rate
\end{itemize}

We can compare alternative models by examining their performance on labeled test data:

\begin{enumerate}
\item Compare {\tt model.predict()} output to the target column:
  \begin{enumerate}
  \item Proportion of predictions that agree.
  \item {\bf True Positive Rate}: out of the people that {\it were} in fact invited, what
    proportion {\it were} predicted to be invited.
  \item True negative rate: out of the people that were {\it not} in fact invited, what proportion
    were {\it not} predicted to be invited.
  \end{enumerate}
\item Compare {\tt model.predict\_proba()} output to the target column:
  \begin{enumerate}
  \item {\bf ``Soft'' True Positive Rate}: out of the people that {\it were} in fact invited, what
    is their mean probability?
  \item Sort by probabilities; verify that highest probabilities were in fact invited.
  \item What is the probability threshold above which predictions are ``accurate''?
  \end{enumerate}
\end{enumerate}



\end{document}
